---
title: "Preparing and searching a PFAS structure and in silico MS/MS library"
output: 
  rmarkdown::html_document:
    toc: True 
    theme: readable
    highlight: tango
---

```{r, include = FALSE}
# output: rmarkdown::html_vignette
# vignette: >
#   %\VignetteIndexEntry{Making the PFAS structure and in silico MS/MS library}
#   %\VignetteEngine{knitr::rmarkdown}
#   %\VignetteEncoding{UTF-8}
# 
# knitr::opts_chunk$set(
#   collapse = TRUE,
#   comment = "#>"
# )
```

## Installation and setup 

```{r, eval = F}
library(devtools)
install_github('gjgetzinger/PFAScreeneR')
install_github('gjgetzinger/cfmR')
```


```{r, eval = T}
suppressMessages({
  library(PFAScreeneR)
  library(reticulate)
  library(DBI)
  library(tidyverse)
  library(cfmR)
  library(MSnbase)
  library(DBI)
  library(dplyr)
  library(eulerr)
  library(stringr)
  library(ggplot2)
  library(tidyr)
  library(ggpubr)
  library(pbapply)
})
```

# Making the molecular database and spectral library 

## Read input molecules, consolidate structures and predict transformations

Input molecules are read from xls files stored in the system files of the `PFAScreeneR` packge. 

```{r, eval = F}
mollist_path <- system.file(package = 'PFAScreeneR') %>% 
  list.dirs() %>% 
  .[grep('mass_lists', .)] %>%
  paste0(., '/')
```

The functions for collecting, standardizing and prediting transformations of input molecules are performed by calling `make_pfas_masslist`, which is imported from a python script stored in the system files of this package. 

```{r, eval = F}
py_file <- system.file(package = 'PFAScreeneR') %>% 
  list.dirs() %>% 
  .[grep('python',.)] %>% 
  list.files(full.names = T)
reticulate::source_python(file = py_file)
```

Biotransformation is performed using the `BioTransformer` algorithm. See  [http://biotransformer.ca](http://biotransformer.ca) for installation intructions and to cite the original work. The path to the `BioTransformer` jar file must be designated. 

```{r}
biotrans_jarfile <- '~/path/to/biotransformerjar/biotransformer-1.1.2.jar'
```

Results are written to SDF and SQLite files. 
```{r}
result_db <- '/path/to/result/PFAScreenR.db'
```

Calling `make_pfas_masslist` performs all the steps for intial database creation. 

```{r, eval = F}
make_pfas_masslist(
  mollist_path = mollist_path, 
  biotrans_jarfile = biotrans_jarfile, 
  hyd_sdf = 'hydrolysis_out.sdf', 
  biotrans_in_sdf = 'biotrans_in.sdf', 
  biotrans_out_sdf = 'biotrans_out.sdf', 
  result_sdf = 'PFAScreeneR.sdf', 
  result_db = result_db
)
```

## Predict MS/MS spectra using CFM 

This step requires execution on a Linux compute cluster running a SLURM scheduler and with CFM-ID installed as a commandline tool. 
See [https://cfmid.wishartlab.com](https://cfmid.wishartlab.com) for installation instructions for CFM-ID. 

```{r, eval = F, echo = T}
conn <- dbConnect(RSQLite::SQLite(), result_db)
mass_list <- tbl(conn, 'molecules') 
```

### Negative ion 

```{r, eval = F}
cfm_predict_batch(
  id = pull(mass_list, ID),
  input_smiles_or_inchi = pull(mass_list, InChI),
  prob_thresh_for_prune = 0.001,
  include_annotations = 1,
  apply_post_processing = 1,
  out_dir = '/path/to/negative_ion/storage/',
  param_filename = '/path/to/negative_ion/param_output1.log',
  config_filename = 'path/to/negative_in/param_config_neg.txt',
  slurm_options = list(partition = 'your_partition', 'mem' = '16G'),
  cpus_per_node = 1,
  nodes = 1000
)
```

### Positive ion 

```{r,eval = F}
cfm_predict_batch(
  id = pull(mass_list, ID),
  input_smiles_or_inchi = pull(mass_list, InChI),
  prob_thresh_for_prune = 0.001,
  include_annotations = 1,
  apply_post_processing = 1,
  out_dir = '/path/to/positive_ion/storage/',
  param_filename = '/path/to/positive_ion/param_output1.log',
  config_filename = '/path/to/positive_ion/param_config_pos.txt',
  slurm_options = list(partition = 'your_partition', 'mem' = '16G'),
  cpus_per_node = 1,
  nodes = 1000
)

```

Disconnect from the result database before attempting to retrieve the results. 

```{r, eval = F}
dbDisconnect(conn)
```

### Collecting results 

Once CFM prediction has completed, results can be written to the databse file. 

```{r, eval = F}
cfm_read_batch(out_dir = '/path/to/negative_ion/storage/',
               table_name = 'cfm_neg',
               output_path = result_db)

cfm_read_batch(out_dir = '/path/to/positive_ion/storage/',
               table_name = 'cfm_pos',
               output_path = result_db)
```

# Searching the spectral library

For demonstration we will use the test, consisting for >300 PFAS MS/MS spectra stored in a MGF file. 

```{r, eval = F}
pfas_exp_msms <-
  readMgfData(list.files(
    system.file(package = 'PFAScreeneR'),
    pattern = '.mgf',
    full.names = T
  ))

pol <-
  ifelse(fData(pfas_exp_msms)$CHARGE == '1-', 'negative', 'positive')
spec_table_name <- ifelse(pol == 'negative', 'cfm_neg', 'cfm_pos')
ID <- as.character(fData(pfas_exp_msms)$ID)
instrument_type <- as.character(fData(pfas_exp_msms)$INSTRUMENT)
ppm <- ifelse(instrument_type == 'Orbitrap', 10, 50)
mol_form <- as.character(fData(pfas_exp_msms)$FORMULA)
exact_mass <-
  as.numeric(as.character(fData(pfas_exp_msms)$EXACT_MASS))
nom_mass_ppm <-
  sapply(exact_mass, function(x)
    1e6 * (diff(x + c(-0.5, 0.5)) / x))
```

## Search by ID

Searching by identifier retrieves library spectra from molecules with the same ID. In this case, the ID assigned to experimental spectra (`pfas_exp_msms`) matches the identifiers used to store molecules in the library. The ID in thise case is the first-14 characters of the hashed InChI string (i.e., InChI Key). 

```{r, eval = F}
sim_id_search <- pbapply::pblapply(seq(along = pfas_exp_msms),
                                   function(x) {
                                     cfm_search(
                                       query_spectrum = pfas_exp_msms[[x]],
                                       spec_table_name = spec_table_name[x],
                                       pol = pol[x],
                                       db_file = db_file,
                                       mol_table_name = 'molecules',
                                       molprop_table_name = 'mol_props',
                                       fun = 'dotproduct',
                                       mol_form = NULL,
                                       exact_mass = NULL,
                                       bin = 0.005,
                                       ID = ID[x],
                                       ppm = ppm[x]
                                     )
                                   })
```

## Search by molecular formula

Searhing by molecular formula retrieves only library spectra with a particular molecular formula. 

```{r, eval = F}
sim_molform_search <- pbapply::pblapply(seq(along = pfas_exp_msms),
                                        function(x) {
                                          cfm_search(
                                            query_spectrum = pfas_exp_msms[[x]],
                                            spec_table_name = spec_table_name[x],
                                            pol = pol[x],
                                            db_file = db_file,
                                            mol_table_name = 'molecules',
                                            molprop_table_name = 'mol_props',
                                            fun = 'dotproduct',
                                            mol_form = mol_form[x],
                                            exact_mass = NULL,
                                            bin = 0.005,
                                            ID = NULL,
                                            ppm = ppm[x]
                                          )
                                        })
```

## Search by neutral exact mass

Searching by neutral exact mass retrieves library spectra with molecular weights within the designated ppm range. 

```{r, eval = F}
sim_neutralmass_search <-
  pbapply::pblapply(seq(along = pfas_exp_msms),
                    function(x) {
                      cfm_search(
                        query_spectrum = pfas_exp_msms[[x]],
                        spec_table_name = spec_table_name[x],
                        pol = pol[x],
                        db_file = db_file,
                        mol_table_name = 'molecules',
                        molprop_table_name = 'mol_props',
                        fun = 'dotproduct',
                        mol_form = NULL,
                        exact_mass = exact_mass[x],
                        bin = 0.005,
                        ID = NULL,
                        ppm = ppm[x]
                      )
                    })
```

## Search by neutral nominal mass

Same as above, except with nominal mass. 

```{r, eval = F}
sim_nominalneutralmass_search <- pbapply::pblapply(seq(along = pfas_exp_msms),
                                                     function(x) {
                                                       cfm_search(
                                                         query_spectrum = pfas_exp_msms[[x]],
                                                         spec_table_name = spec_table_name[x],
                                                         pol = pol[x],
                                                         db_file = db_file,
                                                         mol_table_name = 'molecules',
                                                         molprop_table_name = 'mol_props',
                                                         fun = 'dotproduct',
                                                         mol_form = NULL,
                                                         exact_mass = exact_mass[x],
                                                         bin = 0.005,
                                                         ID = NULL,
                                                         ppm = nom_mass_ppm[x]
                                                       )
                                                     })
```

## Search by exact precursor m/z

Searching by precursor m/z takes the precursor ion m/z and calculates a series of netural masses based on common ESI/APCI adduct ions. The resultant neutral masses are used as described above for neutral mass look-up. 

```{r, eval = F}
sim_precurmz_search <- pbapply::pblapply(seq(along = pfas_exp_msms),
                                         function(x) {
                                           cfm_search(
                                             query_spectrum = pfas_exp_msms[[x]],
                                             spec_table_name = spec_table_name[x],
                                             pol = pol[x],
                                             db_file = db_file,
                                             mol_table_name = 'molecules',
                                             molprop_table_name = 'mol_props',
                                             fun = 'dotproduct',
                                             mol_form = NULL,
                                             exact_mass = NULL,
                                             bin = 0.005,
                                             ID = NULL,
                                             ppm = ppm[x]
                                           )
                                         })
```

## Search by nominal precursor m/z

Same as above, except with nominal mass. 

```{r, eval = F}
sim_nominalprecurmz_search <-
  pbapply::pblapply(seq(along = pfas_exp_msms),
                    function(x) {
                      cfm_search(
                        query_spectrum = pfas_exp_msms[[x]],
                        spec_table_name = spec_table_name[x],
                        pol = pol[x],
                        db_file = db_file,
                        mol_table_name = 'molecules',
                        molprop_table_name = 'mol_props',
                        fun = 'dotproduct',
                        mol_form = NULL,
                        exact_mass = NULL,
                        bin = 0.005,
                        ID = NULL,
                        ppm = nom_mass_ppm[x]
                      )
                    })
```

